A weakly structured stem for human origins in Africa

### 2023-02-20 paper.tex
an attempt to condense the abstract

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here), Mende (from Sierra Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendants of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}).
\aprcomment{See Supplementary Information (SI) section 1.2 for details of
sample collection and consent.}
The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:diversity}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged $550$ka from a common
stem. We computed one- and two-locus statistics whose expectation within and across 
populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.

>>>>

\begin{abstract}
    \aprcomment{An attempt at condensing a bit -- this is just under 230 words}

    Despite broad agreement that \emph{Homo sapiens} originated within Africa,
    considerable uncertainty surrounds specific models of divergence and
    migration across the continent. Progress is hampered by limited fossil and
    genomic data, as well as variability in prior divergence time estimates.
    We use linkage disequilibrium and diversity-based statistics, optimized for
    rapid, complex demographic inference to discriminate among such models.  We
    infer detailed demographic models for populations across Africa, including
    representatives from eastern and western groups and newly sequenced whole
    genomes from 44 Nama (Khoe-San) individuals. We infer a reticulated African
    population history in which contemporary population structure dates back to
    Marine Isotope Stage (MIS) 5. The earliest population divergence among
    contemporary populations occurs 120--135ka and was preceded by structure
    between two or more weakly differentiated ancestral \emph{Homo} populations
    connected by gene flow over hundreds of thousands of years. Such weakly
    structured stem models provide more likely explanations of patterns of
    polymorphism that were previously attributed to contributions from archaic
    hominins in Africa. In contrast to models with archaic introgression, we
    predict fossil remains from coexisting ancestral populations should be
    genetically and morphologically similar, and an inferred 1--4\% of genetic
    differentiation among contemporary human populations can be attributed to
    genetic drift between stem populations. We show that model misspecification
    explains variation in previous divergence time estimates and argue that
    studying a suite of models is key to robust inferences about deep history.

\end{abstract}

@@

We inferred detailed demographic histories using 4x-8x whole-genome sequencing
data for four diverse African populations, comprising the Nama (Khoe-San from
South Africa, newly presented here, see Supplementary Information (SI) section
1.2 for details of
sample collection and consent), Mende (from Sierra Leone, MSL from the
Phase 3 1000 Genomes Project \citep{1000_Genomes_Project_Consortium2015-zq}),
Gumuz (recent descendants of a hunter-gatherer group from Ethiopia
\citep{Gurdasani2015-qy,Gopalan2022-pw}), and eastern African agriculturalists
(Amhara and Oromo from Ethiopia \citep{Gurdasani2015-qy}). The Amhara and Oromo
populations, despite speaking distinct Afro-Asiatic languages, are highly
genetically similar \citep{Pagani2015-pz,Gopalan2022-pw} and thus the two
groups were combined for a larger sample size (Figure~\ref{fig:diversity}). We also
included the British (GBR) from the 1000 Genomes Project in our demographic
models as a representative source of back-to-Africa gene flow and recent
colonial admixture in South Africa. Finally, we used a high-coverage ancient
Neanderthal genome from Vindija Cave, Croatia \citep{Prufer2017-kk} to account
for gene flow from Neanderthals into non-Africans and gauge the relative
time depth of divergence, assuming Neanderthals diverged $550$ka from a common
stem. We computed one- and two-locus statistics whose expectation within and across 
populations can be computed efficiently and that are well suited for both low- and
high-coverage genomes \citep{Ragsdale2019-nt,Ragsdale2020-nz}. Using a
maximum-likelihood inference framework, we then fit to these statistics a
family of parameterized demographic models that involve population splits, size
changes, continuous and variable migration rates, and punctuated admixture
events, to learn about the nature of population structure over the past million
years.
